---
output: github_document
title: "Summary Tables with 'tab'"
author: "Dane Van Domelen <br> vandomed@gmail.com"
date: "`r Sys.Date()`"
bibliography: vignettes/bibliography.bib
always_allow_html: true
nocite: | 
  @gee, 
  @Kmisc, 
  @knitr1,
  @knitr2,
  @knitr3,
  @survey, 
  @survey-paper, 
  @survival-package, 
  @survival-book
  
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  message = FALSE
)

# Load packages
library("tab")
library("knitr")
library("gee")

# Set xtable options
options("xtable.caption.placement" = "top", 
        "xtable.include.rownames" = TRUE, 
        "xtable.comment" = FALSE)
```

## Installation

You can install and load **tab** from GitHub via the following code:

```{r, eval = FALSE}
devtools::install_github("vandomed/tab")
library("tab")
```

## Functions

The purpose of **tab** is to make it easier to create tables for papers, including Table 1's showing characteristics of the sample and summary tables for fitted regression models. Currently, the following functions are included:

* `tabmeans` compares means in two or more groups.
* `tabmedians` compares medians in two or more groups.
* `tabfreq` compares frequencies in two or more groups. 
* `tabmulti` compares multiple variables in two or more groups.
* `tabmeans.svy`, `tabmedians.svy`, `tabfreq.svy`, and `tabmulti.svy` serve the same purposes as the above functions, but for complex survey data.
* `tabglm` summarizes generalized linear models (GLM's) fit via `glm` or `survey::svyglm`. 
* `tabgee` summarizes generalized estimating equation models (GEE's) fit via `gee::gee`. 
* `tabcoxph` summarizes Cox Proportional Hazards models fit via `survival::coxph` or `survey::svycoxph`.

## Table 1's

<!-- A toy dataset called `tabdata` is included in **tab** package. It is a data frame with 15 variables and 300 observations. Let's take a look: -->

<!-- ```{r toydata} -->
<!-- library("tab") -->
<!-- head(tabdata) -->
<!-- ``` -->

You can use `tabmulti` to compare characteristics across levels of a factor variable, e.g. here comparing age, sex, and race by treatment group in the toy dataset `tabdata`.

```{r, results="asis"}
tabmulti(Age + Sex + Race ~ Group, data = tabdata) %>% kable(format = "markdown")
```

To illustrate some options, we can request `Age` and `Race` to print as `Age (years)` and `Race/ethnicity`, compare medians rather than means for age, and include the sample sizes in the column headings:

```{r}
tabmulti(Age + Sex + Race ~ Group, data = tabdata,
         yvarlabels = list(Age = "Age (years)", Race = "Race/ethnicity"),
         ymeasures = c("median", "freq", "freq"),
         listwise.deletion = TRUE,
         n.headings = TRUE) %>% kable(format = "html")
```


## Regression tables

#### GLM's

Logistic regression for 1-year mortality vs. age, sex, and treatment, with the 
binary factor variables displayed in a "compressed" format:

```{r}
fit <- glm(death_1yr ~ Age + Sex + Group, data = tabdata, family = binomial)
fit %>% tabglm(factor.compression = 5) %>% kable(format = "html")
```

#### GEE's

GEE for high blood pressure (measured at 3 time points longitudinally) vs. 
various predictors, with some higher-order terms:

```{r echo = TRUE, eval = FALSE}
tabdata2 <- reshape(data = tabdata,
                    varying = c("bp.1", "bp.2", "bp.3", "highbp.1", "highbp.2", "highbp.3"),
                    timevar = "bp.visit", direction = "long")
tabdata2 <- tabdata2[order(tabdata2$id), ]
fit <- gee(highbp ~ poly(Age, 2, raw = TRUE) + Sex + Race + Race*Sex,
           id = id, data = tabdata2, family = "binomial", corstr = "unstructured")
fit %>% tabgee(data = tabdata2) %>% kable(format = "html")
```

```{r include = FALSE}
tabdata2 <- reshape(data = tabdata,
                    varying = c("bp.1", "bp.2", "bp.3", "highbp.1", "highbp.2", "highbp.3"),
                    timevar = "bp.visit", direction = "long")
tabdata2 <- tabdata2[order(tabdata2$id), ]
fit <- gee(highbp ~ poly(Age, 2, raw = TRUE) + Sex + Race + Race*Sex,
           id = id, data = tabdata2, family = "binomial", corstr = "unstructured")
```

```{r echo = FALSE}
fit %>% tabgee(data = tabdata2) %>% kable(format = "html")
```

Note that we had to set `data = tabdata2` here, because `gee` objects don't store all of the information on factor variables (unlike `glm` objects).


#### Cox proportional hazards

Survival model for mortality vs. predictors, again compressing the factor variables, and requesting slightly differnet columns (i.e. no p-values):

```{r}
library("survival")
fit <- coxph(Surv(time = time, event = delta) ~ Age + Sex + Group, data = tabdata)
fit %>% tabcoxph(factor.compression = 5, columns = c("beta", "hr.ci")) %>% kable(format = "html")
```


## Complex survey data

The functions in **tab** can also accommodate complex survey data. To illustrate with the included dataset `tabsvydata` (which is data from NHANES 2003-2004, except for the made-up variables `time` and `event`), here's a Table 1:

```{r message=FALSE, warning=FALSE}
library("survey")
design <- svydesign(
  data = tabsvydata,
  ids = ~sdmvpsu,
  strata = ~sdmvstra,
  weights = ~wtmec2yr,
  nest = TRUE
)
tabmulti.svy(Age + Race + BMI ~ Sex, design = design) %>% kable(format = "html")
```

And here's a linear regression:

```{r message=FALSE, warning=FALSE}
fit <- svyglm(BMI ~ Age + Sex + Race, design = design)
fit %>% tabglm(factor.compression = 3) %>% kable(format = "html")
```


## Exporting tables, e.g. to Word

All of the functions in **tab** have an argument called `print.html` which can 
be used to export tables to word processors. Setting `print.html = TRUE` will 
result in a HTML table being output to your current working directory. You can 
open the table (e.g. in Chrome) and copy/paste into your report.


## Options for printing in R

I used **knitr**'s `kable` function for the examples here, but other approaches 
should also work (e.g. **xtable**'s `xtable` or **pandoc**'s `pandoc.table`).


## References
